---
title: "Tutorial: Dashboards in Shiny"
subtitle: "rstudio::conf 2017"
author: "Winston Chang and Joe Cheng"
date: "January 13, 2017"
always_allow_html: yes
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
    mathjax: null
    chakra: "libs/remark-latest.min.js"
    css:
      - "default"
      - "css/local-fonts.css"
      - "css/extras.css"
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

## Before we begin

Make sure you get an RStudio Server Pro account and log in!

http://rstudio2017.joecheng.com/signup

_(Supported browsers: Chrome, Firefox, Safari, IE10+. Sorry, no MS Edge.)_

---
layout: true

<footer>If you just walked in: <a href="http://rstudio2017.joecheng.com/signup">http://rstudio2017.joecheng.com/signup</a></footer>

---

## Say hello

#### Please take a moment to introduce yourself to your neighbors.

- Name and affiliation
- How much experience with R
- How much experience with Shiny
- How much experience building dashboards

Over the course of the next two hours, we will be doing a bunch of coding exercises. Please help each other!

```{r echo=FALSE}
countdown::countdown_timer(180)
```

---

## What sets dashboards apart from other apps?

- Automatic updating
- Potentially many viewers looking at the same data
- May or may not be interactive
- "Ten-foot" user interface

---

## Today we're going to cover

- Server: Reactivity concepts that are general, but particularly helpful for dashboarding (Joe Cheng)
- UI: Styling your app to look like a traditional dashboard (Winston Chang)

---
class: middle

# Server

---

## Problem statement

- New data comes in, either continuously or on a schedule (whether every 15s or every 24h)
  - All active sessions should automatically receive the new data
--

- May need to be transformed, aggregated, summarized, etc.
--

- Visualize data as
  - plots
  - tables
  - scalar values (e.g. "Total unique visitors today: 45,124")
--

- Call attention to exceptional conditions (e.g. "Maximum capacity exceeded", "Monthly sales goal reached")

---

## Techniques we'll cover

- Using reactiveFileReader/reactivePoll functions to automatically pick up new data
- Creating networks of reactive expressions
- Optimizing performance

---
class: middle

## reactiveFileReader and reactivePoll

---

## User-driven events vs. data-driven events

In most Shiny apps, reactivity is driven by **user gestures** (mouse clicks, keystrokes).

In dashboards, we want to drive reactivity using **changing data sources** (new database records, file updates).

--

Shiny comes with two functions designed to make it easy to treat changing data as reactive sources:

- `reactiveFileReader`
- `reactivePoll`

---

## But wait...

Can't we just do this?

```{r eval=FALSE}
dataset <- reactive({
  result <- read.csv("data.csv")
{{ invalidateLater(5000)}}
  result
})

output$plot <- renderPlot({
  plot(dataset()) # or whatever
})
```

--

#### This works, but is incredibly wasteful!

Every 5 seconds we will read all the data, and replot it. If `data.csv` changes every 15 minutes on average, then 99.4% of that work will be wasted effort.

Ideally, we only want `dataset` to invalidate when the file actually changes.

---

## reactiveFileReader

- Reads the given file (`"data.csv"`) using the given function (`read.csv`)
- Periodically (every 1000ms?) reads the last-modified time of the file
- If the timestamp changes, then (and only then) re-reads the file

`reactiveFileReader` wraps all of this logic up for you in an easy-to-use reactive expression object.

---
class: middle

## Demo: tutorial_demos/01_file_reader.R

---

## reactiveFileReader

Creates a reactive expression that always has the "up to date" version of your file data.

```{r eval=FALSE}
dataset <- reactiveFileReader(
  intervalMillis = 1000,
  session = session,
  filePath = "data.csv",
  readFunc = read.csv
)

output$plot <- renderPlot({
  plot(dataset()) # or whatever
})
```

---

## reactiveFileReader

Creates a reactive expression that always has the "up to date" version of your file data.

```{r eval=FALSE}
dataset <- reactiveFileReader(
{{ intervalMillis = 1000,}}
  session = session,
  filePath = "data.csv",
  readFunc = read.csv
)
```

_How often to check if the file has changed (based on last-modified time)._

---

## reactiveFileReader

Creates a reactive expression that always has the "up to date" version of your file data.

```{r eval=FALSE}
dataset <- reactiveFileReader(
  intervalMillis = 1000,
{{ session = session,}}
  filePath = "data.csv",
  readFunc = read.csv
)
```

_The user session, or `NULL`._

_If non-null, then the reactive file reader will automatically stop when its session ends._

---

## reactiveFileReader

Creates a reactive expression that always has the "up to date" version of your file data.

```{r eval=FALSE}
dataset <- reactiveFileReader(
  intervalMillis = 1000,
  session = session,
{{ filePath = "data.csv",}}
  readFunc = read.csv
)
```

_The path to the data file. Must be a single file, not a directory name or glob pattern._

---

## reactiveFileReader

Creates a reactive expression that always has the "up to date" version of your file data.

```{r eval=FALSE}
dataset <- reactiveFileReader(
  intervalMillis = 1000,
  session = session,
  filePath = "data.csv",
{{ readFunc = read.csv}}
)
```

_The function that should be called (with `filePath` as an argument) whenever the file changes. The reactive expression then assumes the function's return value._

_We're using a function from `base` here, but you can use any function, including [anonymous functions](http://adv-r.had.co.nz/Functional-programming.html#anonymous-functions)._

---

## reactiveFileReader

Creates a reactive expression that always has the "up to date" version of your file data.

```{r eval=FALSE}
dataset <- reactiveFileReader(
  intervalMillis = 1000,
  session = session,
  filePath = "data.csv",
  readFunc = read.csv,
{{ stringsAsFactors = FALSE}}
)
```

_You can also add any other named arguments, and they'll be passed to `readFunc` whenever it's invoked._

---

## reactivePoll

**reactiveFileReader is limited to files on disk**. It doesn't work for non-file-based data sources like databases or APIs.

--

**reactivePoll is a generalization of reactiveFileReader.** You give it two functions.

---

## reactivePoll

Creates a reactive expression that always has the "up to date" version of your ~~file~~ data.

```{r eval=FALSE}
dataset <- reactivePoll(
  intervalMillis = 1000,
  session = session,
{{ checkFunc = ...,}}
  valueFunc = ...
)
```

_A function that can execute quickly, and merely determine if anything has changed._

---

## reactivePoll

Creates a reactive expression that always has the "up to date" version of your ~~file~~ data.

```{r eval=FALSE}
dataset <- reactivePoll(
  intervalMillis = 1000,
  session = session,
  checkFunc = ...,
{{ valueFunc = ...}}
)
```

_A function with the (potentially expensive) logic for actually reading the data._

---

## Example: Database

The "check" function might return the number of rows in a table, or the max value of a timestamp column.

The "value" function actually returns the contents of the table (or some aggregation thereof).

```{r eval=FALSE}
dataset <- reactivePoll(15000, session,
  checkFunc = function() {
    # Return the latest timestamp
    DBI::dbGetQuery(dbConn, "SELECT MAX(updated) FROM log;")
  },
  valueFunc = function() {
    # Return the 20 most recently updated rows
    DBI::dbGetQuery(dbConn,
      "SELECT * FROM log ORDER BY updated DESC LIMIT 20;"
    )
  }
)
```

---

## Example: Let's reimplement reactiveFileReader

```{r eval=FALSE}
myReader <- function(millis, session, filePath, readFunc, ...) {
  reactivePoll(millis, session,
    checkFunc = function() {
      file.info(filePath)$mtime
    },
    valueFunc = function() {
      readFunc(filePath, ...)
    }
  )
}
```

---

## checkFunc details

**It is important that the `checkFunc` function actually be fast!** It will block the R process while it runs!

(The slower it is, the greater you should make the polling interval.)

--
<br/><br/>

**The `checkFunc` function should not return `TRUE` or `FALSE` for changed/unchanged.**

Instead, just **return a value** (like the timestamp, or the count); it's reactivePoll's job, not yours, to keep track of whether that value is the same as the previous value or not.

---

## Example: Web API

The `checkFunc` could do a partial HTTP request (called a `HEAD` request) that doesn't download actual data, but does tell us whether the data has changed since we last looked (using `Last-Modified` and/or `ETag` HTTP headers):

```{r eval=FALSE}
checkFunc = function() {
  httr::HEAD(apiUrl)$headers[c("last-modified", "etag")]
}
```

The `valueFunc` function does a full HTTP GET request:

```{r eval=FALSE}
valueFunc <- function() {
  httr::GET(apiUrl)
}
```

---

## Your turn

Open `tutorial_exercises/reactive_poll.R`.

This app downloads an RSS feed and summarizes the results in a table. To refresh the results, you have to reload the page. (Ugh, so 2003!)

Modify this app using `reactivePoll` so that the results are always fresh.

```{r echo=FALSE}
countdown::countdown_timer(7 * 60)
```

---

## Solution

Open `tutorial_exercises/reactive_poll.solution.R`.

---
class: middle

## Processing data

---

## Processing data

Besides retrieving data, we might want to process it in other ways.

- Filter
- Summarize
- Detect trends
- Model fitting

How can we perform these calculations against auto-updating data?

---

## Your turn

Open `tutorial_exercises/processing.R`.

Run this app; it shows an auto-updating table of fake stock data. It also shows the symbol and price of the stocks with the highest and lowest share prices.

#### Assignment

Add two columns to the table output:

- `change`     (price change since previous day's closing, in $)
- `change_pct` (price change since previous day's closing, in %)

Also, showing the stocks with highest and lowest share price
is pretty useless. Make those fields show the biggest gainer and loser,
by `change_pct`, instead.

```{r echo=FALSE}
countdown::countdown_timer(10 * 60)
```

---

## Solution

Open `tutorial_exercises/processing.solution.R`.

I solved this by introducing a reactive expression called `full_data`, which adds the desired columns. This guarantees that the `change` and `change_pct` calculations need only be performed once, no matter how may outputs are using it.

```{r eval=FALSE}
raw_data <- reactivePoll(...)

full_data <- reactive({
  df <- raw_data() %>% mutate(
    change = price - prev,
    change_pct = (price - prev) / prev * 100
  )
})

output$all <- renderTable({
  full_data()
})
```

---

## Anti-solution 1

```{r eval=FALSE}
output$all <- renderTable({
  raw_data() %>% mutate(
    change = price - prev,
    change_pct = (price - prev) / prev * 100
  )
})

output$max_gainer <- renderText({
  max_gainer <- raw_data() %>% mutate(
    change = price - prev,
    change_pct = (price - prev) / prev * 100
  ) %>% top_n(1, change_pct)
  paste(max_gainer$symbol, formatC(max_gainer$change_pct))
})
```

---

## Anti-solution 1

```{r eval=FALSE}
output$all <- renderTable({
  raw_data() %>% mutate(
{{   change}} = price - prev,
{{   change_pct}} = (price - prev) / prev * 100
  )
})

output$max_gainer <- renderText({
  max_gainer <- raw_data() %>% mutate(
{{   change}} = price - prev,
{{   change_pct}} = (price - prev) / prev * 100
  ) %>% top_n(1, change_pct)
  paste(max_gainer$symbol, formatC(max_gainer$change_pct))
})
```

Duplicate code! Bad for performance, readability, and maintainability.

---

## Anti-solution 2

```{r eval=FALSE}
values <- reactiveValues(full_data = NULL)

observeEvent(raw_data(), {
  values$full_data <- raw_data() %>% mutate(
    change = price - prev,
    change_pct = (price - prev) / prev * 100
  )
})

output$all <- renderTable({
  values$full_data
})
```

The solution is _reactive_, this anti-solution is _imperative_; more intuitive for beginners to Shiny and reactive programming, but less efficient and impossible to reason about as apps get more complicated.

See the videos on _Effective Reactive Programming_ from [ShinyConf 2016](https://www.rstudio.com/resources/webinars/shiny-developer-conference/) for much more.

---

## Takeaway

Use reactive expressions (`reactive({...})`) to store calculations that need to be reused, as you would introduce a variable in a regular R script.

---

## Performance

Options to improve performance and responsiveness:

1. Cache results
2. Speed up processing logic (Rcpp, parallel)
3. Scale hardware up/out
4. Remove functionality

We'll focus on 1 today.

---

## Example: Performance

`daily-job.R`
```{r eval=FALSE}
{{A}}
```

`app.R`
```{r eval=FALSE}
library(shiny)
{{B}}

ui <- ...

server <- function() {
{{ C}}
  output$plot <- renderPlot({
{{   D}}
  })
}
```

---

## Example: Performance

<table class="tabular">
<tr><th></th><th>Description</th><th>How many</th></tr>
<tr><th>A</th><td>Data processing script</td><td>Once daily</td></tr>
<tr><th>B</th><td>Shiny R process</td><td>4</td></tr>
<tr><th>C</th><td>Shiny session</td><td>100 per process</td></tr>
<tr><th>D</th><td>renderPlot</td><td>5 times per process</td></tr>
</table>

---

## Example: Performance

`daily-job.R`
```{r eval=FALSE}
{{1X}}
```

`app.R`
```{r eval=FALSE}
library(shiny)
{{4X}}

ui <- ...

server <- function() {
{{ 400X}}
  output$plot <- renderPlot({
{{   2,000X}}
  })
}
```

---

## Takeaway

Think carefully about how often your code needs to run, and move it to the scope that matches.

--

Especially think about how to move your code out of your Shiny app and into preprocessing scripts, so you're not in a user's critical path.
